A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-09 podc_paper/boundedspace.tex
working on bounds for GC; discovered problem with using last array inside each node

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m-1$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices smaller than $m-1$ from a RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

\here{Have to talk about repeating CAS to do GC at each node up to p times}

@@

Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
There are at least $p^2$ operations between two GC phases,  so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

>>>>

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
To determine which blocks can be thrown away, the node has an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the RBT 
with indices smaller than $m-1$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices smaller than $m-1$ from a RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.

@@

Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{sec::GCanalysis} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(G)$ steps is spent on a GC phase.
There are at least $G$ operations between two GC phases,  so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-09 podc_paper/correctness-2.tex
working on bounds for GC; discovered problem with using last array inside each node

\subsection{Analysis}

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains $O(q_{max}+p)$ blocks.
\end{lemma}
\begin{proof}
\here{to be written}
\end{proof}

>>>>

\subsection{Analysis}
\label{sec::GCanalysis}

\here{Danger:  what if I remove a block (by a split), which prevents another process from continuing on to mark ancestors/descendants of this node?  Does existing helping take care of this?  Or maybe this does not cause too many excess blocks to be left around?  If not, maybe before splitting, we should mark descendants/ancesotrs of max marked block; in this case, all marking should go in the same direction up/down tree.}

@@

We say the index $b$ \emph{is marked in node $v$} if some process has tested
whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
Intuitively, a block's index value is marked 
when a dequeue passes through this block to find the enqueue
whose argument it should return, or when a dequeue that returns null passes through the block.
Thus, $b$ gets marked in $v$ when $v.\fld{blocks}[b]$ 
contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
If the index of block $B$ is marked in $v$, we also say $B$ is marked in $v$.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains $O(q_{max}+p)$ blocks.
\end{lemma}
\begin{proof}
Consider the \op{AddBlock} that updates a node $v$'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
That \op{AddBlock} performs a GC phase.
Let $C$ be the configuration after it executes line \ref{beforeLastScan}.
Let $L'$ be the prefix of the linearization order $L$ consisting of blocks that have beed added to the
root before $C$.
$A$'s \op{Split} on line \ref{splitGC} removes from the RBT all blocks whose indices
were marked in $v$ prior to line \ref{beforeLastScan}, except for $m$ and $m-1$.
To bound the number of blocks left in the RBT after the split, 
we get a bound on the number of operations in unmarked blocks in $v$.

First consider an enqueue operation $E$ in an unmarked block in $v$.
No dequeue that returns $E$'s value has completed (otherwise it would have marked the block in $v$).
So, if the dequeue is in $L'$, then the dequeue is still pending at $C$.
There can be at most $p$ dequeues pending at $C$.
So, either $E$'s element is still in the

An enqueue operation can be in an unmarked block in $v$
only if the dequeue that returns its value has not yet completed.

\here{to be continued}
\end{proof}
